---
title: "Life Expectancy in Africa"
output: html_document
---

# Using the gapminder data to analyse life expectancy over time in Africa

```{r include=FALSE}
# install.packages("tidyverse")
# install.packages("lubridate")
# install.packages("gapminder")
# library(lubridate)
library(gapminder)
library(dplyr)
library(ggplot2)
```

Examine the dataset
```{r}
head(gapminder)
dim(gapminder)
colnames(gapminder)
unique(gapminder$country)
```

Plot life expectancy in Africa over time
```{r}
africa_exp <- gapminder %>%
   filter(continent=="Africa")

ggplot(africa_exp, aes(x=year, y=lifeExp, color=country ))+
      geom_line()
```
This is too much, maybe we should plot by region, but there is no region data here.

Region data can be imported from another of my repos.
```{bash}
wget https://raw.githubusercontent.com/OneEsk19/Coronavirus/main/CountryList.csv
```
Read this data in
```{r}
geographic <- read.csv("CountryList.csv", header = T)
colnames(geographic) <- c("country", "ISO.alpha3.Code", "Region", "Continent" )
```

Merge the two data sources
```{r}
all_data <- merge(gapminder, geographic, by="country")
all_data <- all_data[,-2]
```

Subset Africa
```{r}
africa_exp <- all_data %>%
      filter(Continent=="Africa")
```
... and plot by region

'facet_grid' plot was tried, but results were confusing as the key was lumped together but the plots were in separate panels.
Below is visually the best outcome, plots are saved to directory

```{r}
for (i in unique(africa_exp$Region)){
      temp_plot = ggplot(africa_exp[africa_exp$Region==i,], aes(x=year, y=lifeExp, color=country )) + geom_line()+ ggtitle(i)
      ggsave(temp_plot, file=paste(i,"_life_exp.png"), width=15, height=11, units="cm")
}
```

To distribute the countries a little more evenly across the plots, I will re-assign some countries that are often considered part of 'Southern Africa' from where there are curretly described as 'Eastern Africa'  
According to this: https://en.wikipedia.org/wiki/Regions_of_the_African_Union#East
There should be the following changes:  
|  Country  |  From  |  To |
____________________________
|  Burundi  |  Eastern  |  Middle   |
|  Mauritania  |  Western  |  North  |
|  Mozambique  |  Eastern  |  Southern  |
|  Malawi  |  Eastern  |  Southern  |
|  Zimbabwe  |  Eastern  |  Southern  |
|  Zambia  |  Eastern  |  Southern  |
|  Angola  |  Eastern  |  Southern  |

```{r}
southern<- c("Mozambique", "Malawi", "Zimbabwe", "Zambia", "Angola")
middle<- "Burundi"
northern <- "Mauritania"

africa_exp <- 
africa_exp %>%
   mutate(Region = replace(Region, (africa_exp$country %in% southern), "Southern Africa")) %>%
   mutate(Region = replace(Region, (africa_exp$country %in% middle), "Middle Africa")) %>%
   mutate(Region = replace(Region, (africa_exp$country %in% northern), "Northern Africa"))
```

Lets also rename 'Middle Africa' to 'Central Africa'
!This was quite trickier than i thought it would be because it was a factor variable and needed special syntax!
```{r}
levels(africa_exp$Region)[levels(africa_exp$Region)=="Middle Africa"] <- "Central Africa"
```

And replot
```{r}
for (i in unique(africa_exp$Region)){
      temp_plot = ggplot(africa_exp[africa_exp$Region==i,], aes(x=year, y=lifeExp, color=country )) + geom_line()+ ggtitle(i)
      ggsave(temp_plot, file=paste(i,"_life_exp.png"), width=15, height=11, units="cm")
}
```
This is better!

Some countries had very obvious drops in life expectancy after 1980, an example of which is Rwanda (Eastern Africa) which presumabl represents the genocide that occurred around this time.  
Looking at this variabililty in life expectancy from this time.  
Subset post 1980
```{r}
post1980 <- 
africa_exp %>%
   filter(year>=1980)
```

Calculate the variance in life expectancy for each country over this time period.
```{r}
life_exp_var_africa <- aggregate(post1980$lifeExp, by=list(post1980$country), FUN=sd)
colnames(life_exp_var_africa) <- c("country", "LE_Variance")


ggplot(life_exp_var_africa, aes(x=country, y=LE_Variance))+
      geom_col()
```
High var only
```{r}
highvar <- 
   life_exp_var_africa %>%
   filter(LE_Variance >= 5) %>% 
   select(country)

highvar_countries <-
   africa_exp %>%
   filter(country %in% highvar$country)
   
   
p <- ggplot(highvar_countries, aes(x=year, y=lifeExp, color=country)) + geom_line() + ggtitle("Life expectancy over time with high variance")

ggsave(p, file="highVar_LE_Africa.png", width=15, height=11, units="cm")
```
It's interesting to see that Egypt, Eritrea, Gambia, and Niger have quite different curves to Botswana, Lesotho, Rwanda, Zambia and Zimbabwe.
While the variance is high for all, the former group has experienced a large and continuos rise in life expectancy, whereas the latter group has experienced a rise, followed by a significant drop, followed by a rise.

Botswana
- Absolutely no idea why the life expectancy dipped from ~ 1990 - 2002
Lesotho
- Lots of leadership disputes and upheavals since early 90's
Rwanada
- genocide which killed up to 1M people in 1994
Zambia
- One party state from 1972-1991 perhaps caused a decline and/or increase in poverty
Zimbabwe
- Economic decline since 90's
- HIV pandemic (This affected all of the Southern African region, and this is reflected in the graphic for this region)
- Authoritarian regimes, wealth redistribution etc